Downstream analyses of whole-genome bisulfite-sequencing data

### 2014-12-06 Rmarkdown/appendix.Rmd
fixed some formatting and Makefile issues

```{r, cache = TRUE}
library(BSgenome.Hsapiens.UCSC.hg19)
library(knitr)
# Nucleotide frequencies
nf <- bsapply(new("BSParams", X = Hsapiens, FUN = letterFrequency,
            exclude = c("rand", "Un", "hap"), simplify = TRUE),
            letters = c("A", "C", "G", "T"))
# Pretty print the results
kable(data.frame(percentage = round(100 * rowSums(nf) /
                                      sum(as.numeric(nf)), 1)))

@@

```{r dinucleotide_distribution, fig.lp = "fig:dinuclotide_distribution", dev = "pdf"}
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
edf <- nf %*% t(nf) / (sum(nf %*% t(nf))) * 100
x <- data.frame(Dinucleotide = names(rowSums(df)),
                Observed = as.vector(100 * rowSums(df) / sum(as.numeric(df))),
                Expected = as.vector(edf))
ggplot(x, aes(x = Dinucleotide, y = Observed / Expected)) +
  geom_point(size = 5) + coord_cartesian(ylim = c(0, 1.5)) +
  ggtitle("CpGs are underrepresented in human reference genome (hg19)") +
  thesis_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 1, linetype = 2)
```

@@

```{r cpg_ipds, cache = TRUE}
library(BSgenome.Hsapiens.UCSC.hg19)
set.seed(666)

@@

```{r cpg_ipds_cdf, dev = "pdf"}
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)

>>>>

```{r, cache = TRUE, message = FALSE}
library(BSgenome.Hsapiens.UCSC.hg19)
library(knitr)
# Nucleotide frequencies
nf <- bsapply(new("BSParams", X = Hsapiens, FUN = letterFrequency,
            exclude = c("rand", "Un", "hap"), simplify = TRUE),
            letters = c("A", "C", "G", "T"))
# Pretty print the results
kable(data.frame(percentage = round(100 * rowSums(nf) /
                                      sum(as.numeric(nf)), 1)))

@@

```{r dinucleotide_distribution, fig.lp = "fig:dinuclotide_distribution", dev = "pdf", message = FALSE}
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
edf <- nf %*% t(nf) / (sum(nf %*% t(nf))) * 100
x <- data.frame(Dinucleotide = names(rowSums(df)),
                Observed = as.vector(100 * rowSums(df) / sum(as.numeric(df))),
                Expected = as.vector(edf))
ggplot(x, aes(x = Dinucleotide, y = Observed / Expected)) +
  geom_point(size = 5) + coord_cartesian(ylim = c(0, 1.5)) +
  ggtitle("CpGs are underrepresented in human reference genome (hg19)") +
  thesis_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 1, linetype = 2)
```

@@

```{r cpg_ipds, cache = FALSE, message = FALSE}
library(BSgenome.Hsapiens.UCSC.hg19)
set.seed(666)

@@

```{r cpg_ipds_cdf, dev = "pdf", message = FALSE}
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)

### 2014-12-06 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
fixed some formatting and Makefile issues

```{r m-bias, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:ADS_M-bias", fig.cap = "M-bias plot for the ADS sample from the Lister (2011) dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.", fig.width = 8, dev = "pdf"}
library(dplyr)
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
read.mbias <- function(file) {

@@

```{r M-bias_spike, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:E18VA_M-bias", fig.cap = "M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. For CpGs in read 1, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in read 2, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read positions for all methylation types. The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.", fig.width = 8, dev = "pdf"}
#files <- paste0("../M-bias/", list.files("../M-bias/"))
# E18VA
files <- "../data/M-bias/E18VA.M-bias.txt"
sample_names <- gsub("../data/M-bias/", "", gsub(".M-bias.txt", "", files))
mbias <- lapply(files, read.mbias)
mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
                      do.call("rbind", mbias)))
mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
mbias %>%
  filter(sample == "E18VA") %>%
  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: E18VA") +
  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
  thesis_theme +
  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
```

@@

```{r cpgs_per_read, fig.lp = "fig:Lister_cpgs_per_read", dev = "pdf", eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "Number of CpGs per read for the Lister data."}
library(dplyr)
library(ggplot2)
library(scales)
thesis_theme <- theme_classic(base_size = 20)
thesis_theme <- theme_bw(base_size = 20)
dataset <- "Lister"
hist <- lapply(paste0("../data/observable_m-tuples/", dataset, "/",
                      list.files(paste0("../data/observable_m-tuples/",
                                        dataset), "\\.CG\\_per\\_read\\.hist")),
               read.table, header = TRUE)
sn <- gsub("\\.CG\\_per\\_read\\.hist", "",
           list.files(paste0("../data/observable_m-tuples/",
                             dataset), "\\.CG\\_per\\_read\\.hist"))
sn <- gsub("DM\\_", "", sn)
hist_df <- bind_rows(hist) %>%
  mutate(Sample = rep(sn, sapply(hist, nrow))) %>%
  group_by(Sample) %>%
  mutate(Proportion = count / sum(as.numeric(count)))
hist_df %>%
  ggplot(aes(x = n, y = Proportion * 100)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Sample, ncol = 5, nrow = 4) +
  coord_cartesian(xlim = c(-1, 8.5)) +
  scale_y_continuous(label = comma) +
  xlab("Number of CpGs per read\n(Axis truncated at 8)") +
  ylab("Percentage of total reads") +
  ggtitle(paste0(dataset, ": Number of CpGs per read")) +
  thesis_theme
```

@@

```{r methtuple_runtime, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_runtime", fig.cap = "The running times are the 'User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix 'ac' on the tuple size means that the option \texttt{--all-combinations} was set.", fig.width = 8, dev = "pdf"}
library(dplyr)
library(ggplot2)
library(stringr)
library(gridExtra)

@@

```{r methtuple_memory, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_memory", fig.cap = "The maximum memory usage is the 'Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix 'ac' means that the option \texttt{--all-combinations} was set.", fig.width = 8, dev = "pdf"}
# Memory
memory <- lapply(paste0("../data/methtuple_performance/memory/",
                        list.files("../data/methtuple_performance/memory/")),
                 read.table)
names(memory) <- gsub(".memory", "",
                      list.files("../data/methtuple_performance/memory/"))
n <- str_split(names(memory), "\\.")
# NB: Need to divide reported memory usage by 4 due to bug in 'time'
# (https://bugzilla.redhat.com/show_bug.cgi?id=703865)
memory <- data_frame("Chromosome" = factor(rep(sapply(n, "[", 1),
                                               times = sapply(memory, nrow)),
                                           levels =
                                             paste0("chr",
                                                    c(1:22, "X", "Y", "M", "L"))),
                     "Tuple" = rep(sapply(n, "[", 2),
                                   times = sapply(memory, nrow)),
                     "Memory" = unlist(memory) / 4)
gg_memory <- memory %>%
  ggplot(aes(x = Chromosome, y = Memory / 10^6)) + geom_boxplot() +
  facet_wrap( ~ Tuple, ncol = 3) + thesis_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylab("Memory usage (GB)") +
  ggtitle("Maximum memory usage of methtuple across 49 samples")
facet_wrap_labeller(gg_memory, paste0("Tuple size: ", unique(memory$Tuple)))
```

>>>>

```{r m-bias, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:ADS_M-bias", fig.cap = "M-bias plot for the ADS sample from the Lister (2011) dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.", dev = "pdf", outwidth = "0.8\\linewidth"}
library(dplyr)
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
read.mbias <- function(file) {

@@

```{r M-bias_spike, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:E18VA_M-bias", fig.cap = "M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. For CpGs in read 1, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in read 2, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read positions for all methylation types. The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.", dev = "pdf", outwidth = "0.8\\linewidth"}
#files <- paste0("../M-bias/", list.files("../M-bias/"))
# E18VA
files <- "../data/M-bias/E18VA.M-bias.txt"
sample_names <- gsub("../data/M-bias/", "", gsub(".M-bias.txt", "", files))
mbias <- lapply(files, read.mbias)
mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
                      do.call("rbind", mbias)))
mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
mbias %>%
  filter(sample == "E18VA") %>%
  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: E18VA") +
  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
  thesis_theme +
  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
```

@@

```{r cpgs_per_read, fig.lp = "fig:Lister_cpgs_per_read", dev = "pdf", eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "Number of CpGs per read for the Lister data.", outwidth = "0.8\\linewidth"}
library(dplyr)
library(ggplot2)
library(scales)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
dataset <- "Lister"
hist <- lapply(paste0("../data/observable_m-tuples/", dataset, "/",
                      list.files(paste0("../data/observable_m-tuples/",
                                        dataset), "\\.CG\\_per\\_read\\.hist")),
               read.table, header = TRUE)
sn <- gsub("\\.CG\\_per\\_read\\.hist", "",
           list.files(paste0("../data/observable_m-tuples/",
                             dataset), "\\.CG\\_per\\_read\\.hist"))
sn <- gsub("DM\\_", "", sn)
hist_df <- bind_rows(hist) %>%
  mutate(Sample = rep(sn, sapply(hist, nrow))) %>%
  group_by(Sample) %>%
  mutate(Proportion = count / sum(as.numeric(count)))
hist_df %>%
  ggplot(aes(x = n, y = Proportion * 100)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Sample, ncol = 5, nrow = 4) +
  coord_cartesian(xlim = c(-1, 8.5)) +
  scale_y_continuous(label = comma) +
  xlab("Number of CpGs per read\n(Axis truncated at 8)") +
  ylab("Percentage of total reads") +
  ggtitle(paste0(dataset, ": Number of CpGs per read")) +
  thesis_theme
```

@@

```{r methtuple_runtime, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_runtime", fig.cap = "The running times are the 'User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix 'ac' on the tuple size means that the option \texttt{--all-combinations} was set.", dev = "pdf", outwidth = "0.8\\linewidth"}
library(dplyr)
library(ggplot2)
library(stringr)
library(gridExtra)

@@

```{r methtuple_memory, eval = TRUE, echo = FALSE, message = FALSE, fig.lp = "fig:methtuple_memory", fig.cap = "The maximum memory usage is the 'Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix 'ac' means that the option \texttt{--all-combinations} was set.", dev = "pdf", outwidth = "0.8\\linewidth"}
# Memory
memory <- lapply(paste0("../data/methtuple_performance/memory/",
                        list.files("../data/methtuple_performance/memory/")),
                 read.table)
names(memory) <- gsub(".memory", "",
                      list.files("../data/methtuple_performance/memory/"))
n <- str_split(names(memory), "\\.")
# NB: Need to divide reported memory usage by 4 due to bug in 'time'
# (https://bugzilla.redhat.com/show_bug.cgi?id=703865)
memory <- data_frame("Chromosome" = factor(rep(sapply(n, "[", 1),
                                               times = sapply(memory, nrow)),
                                           levels =
                                             paste0("chr",
                                                    c(1:22, "X", "Y", "M", "L"))),
                     "Tuple" = rep(sapply(n, "[", 2),
                                   times = sapply(memory, nrow)),
                     "Memory" = unlist(memory) / 4)
gg_memory <- memory %>%
  ggplot(aes(x = Chromosome, y = Memory / 10^6)) + geom_boxplot() +
  facet_wrap( ~ Tuple, ncol = 3) + thesis_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylab("Memory usage (GB)") +
  ggtitle("Maximum memory usage of methtuple across 49 samples")
facet_wrap_labeller(gg_memory, paste0("Tuple size: ", unique(memory$Tuple)))
```

### 2014-12-06 Rmarkdown/wgbs_downstream_analyses.Rmd
fixed some formatting and Makefile issues

Such a contingency table can be analysed in many different ways. Mathematically, to test whether this cytosine is a DMC we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j. $\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$).

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/MethPat.pdf}
\caption{Schematic of the \texttt{MethPat} class, shown here for 3-tuples. Each row represents a 3-tuple to which the genomic co-ordinates of the tuples (green box) and the frequencies of the methylation patterns (grey box) are aligned. The frequencies of each methylation pattern --- $MMM, MMU, \ldots, UUU$ --- are stored as separate matrices where the columns represent samples --- $S_{1}, \ldots, $S_{n}$. Some sample may have now have any sequencing coverage for a particular m-tuple, in which case the corresponding frequencies are recorded as \texttt{NA}.}
\label{fig:MethPat}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lllll@{}}
\toprule
& \texttt\{pryr::object\_size(x)\} & \texttt\{nrow\} & Number of assays & Percentage of \texttt\{NA\} and $0$ values \\ \midrule
1-tuples                                 & $5.9$ GB                        & $56,348,522$    & $2$              & $28\%$                                     \\
2-tuples                                 & $20.1$ GB                        & $100,586,237$   & $4$              & $80\%$                                     \\
2-tuples (\texttt\{--all-combinations\}) & $60.0$ GB                          & $299,814,999$   & $4$              & $78\%$                                     \\
3-tuples                                 & $43.3$ GB                        & $109,376,348$   & $8$              & $93\%$                                     \\
4-tuples                                 & $80.5$ GB                        & $102,625,758$   & $16$             &                                            \\ \bottomrule
\end{tabular}
\caption{Size of \texttt{MethPat} objects for the \em{EPISCOPE} data. All m-tuples are stranded. The size of the \code{MethPat} object is computed using the \texttt{object_size} function in the \texttt{pryr} package (\url{http://cran.r-project.org/web/packages/pryr/index.html}). The number of rows (\texttt{nrow}) corresponds to the number of m-tuples in the object. The number of assays is $2^{\text{m}}$, where $m$ is the size of the m-tuples. The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an `NA` value means that that m-tuple was not observed in that particular sample.}
\label{tab:methpat_sparsity}
\end{table}

>>>>

Such a contingency table can be analysed in many different ways. Mathematically, to test whether this cytosine is a DMC we wish to test the null hypothesis  $H_{0}: \beta_{i, j} = \beta_{i, 0}$ for $j = 1, \ldots, n$ against the alternative that $H_{A}: \beta_{i, j} \neq \beta_{i, 0}$ for some $j$. $\beta_{i, 0}$ may be pre-specified or estimated from the data by, for example, the group average $\frac{1}{n} \sum_{j = 1}^{n} \beta_{i, j}$).

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/MethPat.pdf}
\caption{Schematic of the \texttt{MethPat} class, shown here for 3-tuples. Each row represents a 3-tuple to which the genomic co-ordinates of the tuples (green box) and the frequencies of the methylation patterns (grey box) are aligned. The frequencies of each methylation pattern --- $MMM, MMU, \ldots, UUU$ --- are stored as separate matrices where the columns represent samples --- $S_{1}, \ldots, S_{n}$. Some sample may have now have any sequencing coverage for a particular m-tuple, in which case the corresponding frequencies are recorded as \texttt{NA}.}
\label{fig:MethPat}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lllll@{}}
\toprule
& \texttt{pryr::object\_size(x)} & \texttt{nrow} & Number of assays & Percentage of \texttt{NA} and $0$ values \\ \midrule
1-tuples                                 & $5.9$ GB                        & $56,348,522$    & $2$              & $28\%$                                     \\
2-tuples                                 & $20.1$ GB                        & $100,586,237$   & $4$              & $80\%$                                     \\
2-tuples (\texttt{--all-combinations}) & $60.0$ GB                          & $299,814,999$   & $4$              & $78\%$                                     \\
3-tuples                                 & $43.3$ GB                        & $109,376,348$   & $8$              & $93\%$                                     \\
4-tuples                                 & $80.5$ GB                        & $102,625,758$   & $16$             &                                            \\ \bottomrule
\end{tabular}
\caption{Size of \texttt{MethPat} objects for the \em{EPISCOPE} data. All m-tuples are stranded. The size of the \texttt{MethPat} object is computed using the \texttt{object\_size} function in the \texttt{pryr} package (\url{http://cran.r-project.org/web/packages/pryr/index.html}). The number of rows (\texttt{nrow}) corresponds to the number of m-tuples in the object. The number of assays is $2^{\text{m}}$, where $m$ is the size of the m-tuples. The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
\label{tab:methpat_sparsity}
\end{table}
